---
title: "Workshop setup instructions"
subtitle: "Workshop: Intermediate R Programming"
output: html_document
---

- Make sure that you have installed the [latest versions of R and RStudio](r_and_rstudio.html). 
- In RStudio, install the [devtools and educer packages](https://educe-ubc.github.io/educer.html). This will install all of the tidyverse packages and other resources.
- Install the lmerTest and roxygen2 packages by entering the following code into the console:
    
  ```{r install-lmer-roxygen, eval = FALSE}
  if (!require(lmerTest)) install.packages("lmerTest")
  if (!require(roxygen2)) install.packages("roxygen2")
  ```

- Click on the button below to download the data file we will use during the workshop and save it on your Desktop or another convenient location.
  
```{r echo = FALSE}
data_file <- read.csv("https://raw.githubusercontent.com/EDUCE-UBC/educer/main/data-raw/data_intermediate_ws.csv")

downloadthis::download_this(data_file,
    output_name = "data_intermediate_R",
    output_extension = ".csv",
    button_label = "Download data",
    button_type = "default",
    has_icon = TRUE,
    icon = "fa fa-save",
    class = "hvr-sweep-to-left",
    csv2 = FALSE
  )
```

  
