---
title: "Workshop setup instructions"
subtitle: "Workshop: Introduction to the R tidyverse"
output: html_document
---

- Make sure that you have installed the [latest versions of R and RStudio](r_and_rstudio.html). 
- In RStudio, install the [educer package](https://educe-ubc.github.io/educer.html). This will install all of the tidyverse packages and other resources.
- If you encountered any unexpected errors while installing the educer package, copy the following code into the Console, after the prompt `>`

    ```{r install-packages, eval = FALSE}
    if (!require(tidyverse)) install.packages("tidyverse")
    if (!require(cowplot)) install.packages("cowplot")
    ```

- Download the data file we will use during the workshop by clicking on the following button and save it in your Desktop or another convenient location.
  
    ```{r echo = FALSE}
    data_file <- read.csv("https://raw.githubusercontent.com/EDUCE-UBC/educer/main/data-raw/data_tidyverse_ws.csv")
    
    downloadthis::download_this(data_file,
        output_name = "data",
        output_extension = ".csv",
        button_label = "Download data",
        button_type = "default",
        has_icon = TRUE,
        icon = "fa fa-save",
        class = "hvr-sweep-to-left",
        csv2 = FALSE
      )
    ```

  